---
title: "Mariage Liciences Pre and Post Pandemic"
subtitle: "Was there a difference? "
author: 
  - Fatimah Yunusa
thanks: "Code and data are available at: https://github.com/fatimahsy/Pandemic-Marriages-.git."
date: today
date-format: long
abstract: "This paper examines trends in marriage licenses in Toronto, focusing on the effects of the COVID-19 pandemic on marital patterns. Using a dataset of marriage license registrations from 2015-2023, we apply time series analysis and statistical comparisons to investigate shifts in marriage rates before, during, and after the pandemic. Our findings show a significant decline in licenses issued during the pandemic, followed by a sharp increase in the post-pandemic period. These results suggest that social disruptions caused by the pandemic have had lasting effects on marriage trends, potentially influencing future policy and service planning ."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/analysis_data"))

```

We use @citeR and @opendata.

# Data {#sec-data}

Some of our data is of Marriage licences (@fig-marriage) from @opendatatoronto. The scatter plot (@fig-marriage) displays the number of marriage licenses issued in Toronto over time, spanning from before 2015 to approximately 2025. The graph shows relatively stable issuance of licenses between 2015 and 2019, with a noticeable drop around 2020, coinciding with the onset of the COVID-19 pandemic. Following this decline, there is a sharp increase in licenses after 2021, suggesting a recovery and potential backlog of marriages post-pandemic. The pronounced spike in recent years indicates a significant shift in marriage behavior, likely influenced by the social and economic impacts of the pandemic.

```{r}
#| label: fig-marriage
#| fig-cap: Marriage Licinces By Date
#| echo: false

data |> 
  ggplot(aes(x = date, y = marriage_licenses)) +
  geom_point()

```

\`\`\`

\newpage

\appendix

# Appendix {.unnumbered}

# References
